Missing the Point\(\colon\) Non-Convergence in Iterative Imputation Procedures

### 2020-04-21 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
New results section (draft)

@@

```{r, include = FALSE}
load("../Results/test.Rdata")
source("../2.CreateFigures.R")
```

```{r}
mean_bias + plot_annotation(title = "Bias in mean estimate, convergence of chain means")
mean_Rh 
mean_ACF + plot_annotation(title = "Dotted lines are old AC computation, solid are new.")
```

@@

```{r}
est_bias + plot_annotation(title = "Bias in regression estimate, coverage rate, \n bias in variance explained")
est_cov
Rsq_bias 
```

@@

```{r include=FALSE}
# chain means
dat[-1, 6:9] %>% mutate(max_Rh = apply(., 1, max)) %>% .$max_Rh<1.2

# chain variances
dat[-1, 26:29] %>% mutate(max_Rh = apply(., 1, max)) %>% .$max_Rh<1.2

>>>>

Research question when and how to diagnose non convergence?

QUESTION: How?

The two challenges of convergence are stationarity and mixing point. Mixing and stationarity can be evaluated visually. Common practice is to evaluate stationarity and mixing in trace plots for scalar summaries of interest. In mice these scalar summaries of interest are chain means and chain variances. As Van Buren describes users can also specify a model specific scalar summary (e.g., a regression coefficient). search a user specified scalar summary however is not universal to all complete data problems. Therefore as inspired by someone 2002 we will use introduce a scalar summary that summarizes the multivariate state space of the algorithm. In this study we use the first eigenvalue of the variance covariance matrix of the completed data set per imputation. The methods under evaluation are Rhat and AC. These will be computed for the three scalar summaries of interest: chain means, chain variances, and the first eigenvalue of the variance-covariance matrix across imputations.

The two methods will be evaluated with an estimand per scalar summary of interest. It is assumed that when the estimates are unbiased (and confidence valid), the algorithm is sufficiently converged. The R hat and AC values with chain means as theta are evaluated against the bias in univariate mean estimates (**just for Y??**). Similarly, the performance measure for Rhat and AC applied to chain variances is the bias in estimated standard deviations. To evaluate the performance of Rhat and AC on the eigenvalues, we will use bias in R squared and in the estimated regression coefficient, the coverage rate of the CI95% of the regression estimate, and the CI length. **remove the term performance measure above and add that performance measures are bias in all estimands, and cov and CIL of regression coeff.**

ANSWER

In general there is more bias in the estimate in conditions with a higher proportion of missingness and or a lower number of iterations. As required the values of or hat R lower for conditions with a higher number off iterations. End somewhat higher 4 conditions with a higher percentage of missing data. However autocorrelation did not show a decrease with increase number of iterations. After replacing the ACF function with a manual calculation for autocorrelation the autocorrelation values were indeed decreasing with a higher number of iterations. Evaluation with the performance measures shows that are had an autocorrelation are conservative: signs of non-convergence in conditions where estimands are unbiased and or confidence valid. As expected complete convergence (as indicated by autocorrelation equals 0; R hat equals 1)  is not observed.

QUESTION: When?

Usually, convergence is diagnosed when Rhat < 1.2 or 1.1 or 1.01. And a t-test is performed for AC values == 0. Performance measures are the same as above: unbiased, confidence valid estimates.

ANSWER

Our head is equal to 1.2 is not stringent enough because the performance measures R not unbiased and confidence fellas fell it 4 every condition in which our hat is smaller than one point 2. At least, not when our hat is smaller than one point two fur first occurs. The threshold R hat is equal to one point one is somewhat conservative with respect to the performance measures. It seems OK. Our hat is equal to 1.01 is too stringent compared to the performance measures (**but may be necessary in more complex complete data models???**).

RECOMMENDATIONS

For empirical researchers: 1) Check trace plots for pathological non convergence and adjust imputation model if necessary. 2) use our hat wait 1.1 ash threshold and autocorrelation with **?** as threshold . Keep iterating until these thresholds are reached. 3) Do not use the R function ACF. Instead compute autocorrelations manually , see for example GitHub link. 4) Track your own scalar summary of interest. This is advanced but explained in Van Buren 2018. Compute R hat and autocorrelation values for this scalar summary.

LIMITATIONS/FUTURE RESEARCH

Pmm,
M(N)AR,
Empirical data,
Vignette,
ShinyMICE,
AC across imputations, not iterations (unique for imp data??)

If I don't include it in my study, also: Significance of AC values; Estimate (PCA or reg. coeff.) versus Rhat and AC on the estimate

## Start methods section

@@

```{r, include = FALSE}
# load("../Results/test.Rdata")
# source("../2.CreateFigures.R")
```

```{r}
# mean_bias + plot_annotation(title = "Bias in mean estimate, convergence of chain means")
# mean_Rh 
# mean_ACF + plot_annotation(title = "Dotted lines are old AC computation, solid are new.")
```

@@

```{r}
# est_bias + plot_annotation(title = "Bias in regression estimate, coverage rate, \n bias in variance explained")
# est_cov
# Rsq_bias 
```

@@

```{r include=FALSE}
# # chain means
# dat[-1, 6:9] %>% mutate(max_Rh = apply(., 1, max)) %>% .$max_Rh<1.2
# 
# # chain variances
# dat[-1, 26:29] %>% mutate(max_Rh = apply(., 1, max)) %>% .$max_Rh<1.2

### 2020-04-21 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
New results section (draft)

\begin{abstract}
(\textbf{Rewrite:}) Multiple imputation by chained equations (mice) is a
widely used tool to accommodate missing data. While empirical evidence
supports the validity of inferences obtained using mice, there is no
consensus on the convergence properties of the method. This paper
provides insight into non-convergence of mice algorithms.
\end{abstract}

@@

Missing data problems are ubiquitous and pervasive in the social and
behavioral sciences. If a dataset contains just one missing value for a
variable of interest, statistical inferences are undefined and will not
produce any result. Incomplete observations are therefore ignored by
default in many statistical packages (i.e., list-wise deletion is
employed). Unfortunately, this \emph{ad hoc} solution may yield wildly
invalid results \citep{buur18}. An alternative is to `impute' (i.e.,
fill in) every missing value in an incomplete dataset. With imputation
techniques, one or several completed datasets are created, on which
statistical inferences can be performed. The case with several completed
datasets is known as `multiple imputation' \citep[MI;][]{rubin76}, and
requires an additional step to pool the results across imputations. By
applying Rubin's rules, the pooled result will reflect the uncertainty
in the data due to missingness. Under many circumstances, this yields an
unbiased and confidence valid estimate of the inference on the true--but
missing--data \citep{buur18}. A popular method to obtain imputations is
to use the `Multiple Imputation by Chained Equations' algorithm,
shorthand `MICE'\citep{mice}. MICE is an iterative algorithmic procedure
to draw imputations from the posterior predictive distribution of the
missing values. This introduces a potential threat to the validity of
the imputations: what if the algorithm has not converged? Are the
implications then to be trusted? And can we rely on the inference
obtained on the completed data? These are all open questions, because
the convergence properties of iterative imputation algorithms have not
been systematically studied \citep{buur18}. Moreover, there is no
scientific consensus on the convergence properties of MI algorithms
\citep{taka17}. Some default MICE techniques (e.g., `predictive mean
modeling') might not yield converged states at all \citep{murr18}.
Therefore, algorithmic convergence should be monitored carefully.

@@

Let \(y\) denote an \(n \times p\) matrix containing the data values on
\(p\) variables for all \(n\) units in a sample. The data value of unit
\(i\) (\(i = 1, \dots, n\)) on variable \(j\) (\(j = 1, \dots, p\)) may
be either observed or missing. The collection of observed data values in
\(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred
to as \(y_{mis}\). For each datapoint in \(y_{mis}\), we sample
\(M \times T\) times plausible values, where \(M\) is the number of
imputations (\(m = 1, \dots, M\)) and \(T\) is the number of iterations
(\(t = 1, \dots, T\)). The collection of samples between the initial
value (at \(t=1\)) and the final imputed value (at \(t=T\)) will be
referred to as an `imputation chain'. \textbf{Add: missingness
percentage = proportion of cases with one or more missing values }

@@

From the between- and within-chain variances we compute a weighted
average, \(\widehat{\operatorname{var}}^{+}\), which over-estimates the
total variance of \(\theta\). \(\widehat{R}\) is then obtained as a
ratio between the over-estimated total variance and the within-chain
variance:

@@

Based on an empirical finding \citep{lace07}, we hypothesize that
\(\widehat{R}\) will over-estimate non-convergence of MI algorithms. The
threshold of \(\widehat{R} < 1.01\) will then be too stringent for
diagnosing convergence. This over-estimation may, however, be diminished
because \(\widehat{R}\) can falsely diagnose convergence if initial
values of the algorithm are not appropriately over-dispersed
\citep[p.~437]{broo98}. In \texttt{mice}, initial values are chosen
randomly from the observed data. Therefore, we cannot be certain that
the initial values are over-dispersed. We expect this to have little
effect on the hypothesized performance of \(\widehat{R}\). \textbf{(Add
actual hypothesis: we thought that mice would converge sooner than
\(\widehat{R}\) would indicate that it did)} No hypothesis was
formulated about the performance of \(AC\) as convergence diagnostic.
\textbf{Added later: High \(AC\)-values are implausible in MI
procedures. That is, the randomness induced by the MI algorithm
effectively mitigates the risk of dependency within chains.}

@@

Convergence of the MICE algorithm is investigated through model-based
simulation in \texttt{R} \citep[version 3.6.3;][]{R}. The simulation
set-up is summarized in the pseudo-code below. The complete \texttt{R}
script of the simulation study is available from
\href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.

@@

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-1} \end{flushleft}

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-2} \end{flushleft}

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-3} \end{flushleft}

@@

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-1} \end{flushleft}

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-2} \end{flushleft}

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-3} \end{flushleft}

>>>>

\begin{abstract}
(\textbf{Rewrite:}) Multiple imputation by chained equations (MICE) is a
widely used tool to accommodate missing data. While empirical evidence
supports the validity of inferences obtained using mice, there is no
consensus on the convergence properties of the method. This paper
provides insight into non-convergence of mice algorithms.
\end{abstract}

@@

Missing data problems are ubiquitous and pervasive in the social and
behavioral sciences. If a dataset contains just one missing value for a
variable of interest, statistical inferences are undefined and will not
produce any result. Incomplete observations are therefore ignored by
default in many statistical packages (i.e., list-wise deletion is
employed). Unfortunately, this \emph{ad hoc} solution may yield wildly
invalid results \citep{buur18}. An alternative is to `impute' (i.e.,
fill in) every missing value in an incomplete dataset. With imputation
techniques, one or several completed datasets are created, on which
statistical inferences can be performed. The case with several completed
datasets is known as `multiple imputation' \citep[MI;][]{rubin76}, and
requires an additional step to pool the results across imputations
\textbf{see figure 1??}. By applying Rubin's rules, the pooled result
will reflect the uncertainty in the data due to missingness. Under many
circumstances, this yields an unbiased and confidence valid estimate of
the inference on the true--but missing--data \citep{buur18}. A popular
method to obtain imputations is to use the `Multiple Imputation by
Chained Equations' algorithm, shorthand `MICE'\citep{mice}. MICE is an
iterative algorithmic procedure to draw imputations from the posterior
predictive distribution of the missing values. This introduces a
potential threat to the validity of the imputations: what if the
algorithm has not converged? Are the implications then to be trusted?
And can we rely on the inference obtained on the completed data? These
are all open questions, because the convergence properties of iterative
imputation algorithms have not been systematically studied
\citep{buur18}. Moreover, there is no scientific consensus on the
convergence properties of MI algorithms \citep{taka17}. Some default
MICE techniques (e.g., `predictive mean modeling') might not yield
converged states at all \citep{murr18}. Therefore, algorithmic
convergence should be monitored carefully.

@@

Let \(y\) denote an \(n \times p\) matrix containing the data values on
\(p\) variables for all \(n\) units in a sample. The data value of unit
\(i\) (\(i = 1, \dots, n\)) on variable \(j\) (\(j = 1, \dots, p\)) may
be either observed or missing. The collection of observed data values in
\(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred
to as \(y_{mis}\). For each datapoint in \(y_{mis}\), we sample
\(M \times T\) times plausible values, where \(M\) is the number of
imputations (\(m = 1, 2, \dots, M\)) and \(T\) is the number of
iterations (\(t = 1, 2, \dots, T\)). The collection of samples between
the initial value (at \(t=1\)) and the final imputed value (at \(t=T\))
will be referred to as an `imputation chain'. \textbf{Add: missingness
percentage = proportion of cases with one or more missing values, Q is
quantity of scientific interest: ``A scientific estimand \(Q\) is a
quantity of scientific interest that we can calculate if we would
observe the entire population'' \citep[par 2.3.1]{buur18}. }

@@

From the between- and within-chain variances we compute a weighted
average, \(\widehat{\operatorname{var}}^{+}\), which over-estimates the
total variance of \(\theta\) \textbf{remove or explain why}.
\(\widehat{R}\) is then obtained as a ratio between the over-estimated
total variance and the within-chain variance:

@@

Based on an empirical finding \citep{lace07}, we hypothesize that
\(\widehat{R}\) will over-estimate non-convergence of MI algorithms
\textbf{explain why?}. The threshold of \(\widehat{R} < 1.01\) will then
be too stringent for diagnosing convergence. This over-estimation may,
however, be diminished because \(\widehat{R}\) can falsely diagnose
convergence if initial values of the algorithm are not appropriately
over-dispersed \citep[p.~437]{broo98}. In \texttt{mice}, initial values
are chosen randomly from the observed data. Therefore, we cannot be
certain that the initial values are over-dispersed. We expect this to
have little effect on the hypothesized performance of \(\widehat{R}\)
\textbf{explain that the randomness induced by the MICE algorithm would
take care of this??}. \textbf{(Add actual hypothesis: we thought that
mice would converge sooner than \(\widehat{R}\) would indicate that it
did)} No hypothesis was formulated about the performance of \(AC\) as
convergence diagnostic. \textbf{Added later: High \(AC\)-values are
implausible in MI procedures. That is, the randomness induced by the MI
algorithm effectively mitigates the risk of dependency within chains.}

@@

Research question when and how to diagnose non convergence?

QUESTION: How?

The two challenges of convergence are stationarity and mixing point.
Mixing and stationarity can be evaluated visually. Common practice is to
evaluate stationarity and mixing in trace plots for scalar summaries of
interest. In mice these scalar summaries of interest are chain means and
chain variances. As Van Buren describes users can also specify a model
specific scalar summary (e.g., a regression coefficient). search a user
specified scalar summary however is not universal to all complete data
problems. Therefore as inspired by someone 2002 we will use introduce a
scalar summary that summarizes the multivariate state space of the
algorithm. In this study we use the first eigenvalue of the variance
covariance matrix of the completed data set per imputation. The methods
under evaluation are Rhat and AC. These will be computed for the three
scalar summaries of interest: chain means, chain variances, and the
first eigenvalue of the variance-covariance matrix across imputations.

The two methods will be evaluated with an estimand per scalar summary of
interest. It is assumed that when the estimates are unbiased (and
confidence valid), the algorithm is sufficiently converged. The R hat
and AC values with chain means as theta are evaluated against the bias
in univariate mean estimates (\textbf{just for Y??}). Similarly, the
performance measure for Rhat and AC applied to chain variances is the
bias in estimated standard deviations. To evaluate the performance of
Rhat and AC on the eigenvalues, we will use bias in R squared and in the
estimated regression coefficient, the coverage rate of the CI95\% of the
regression estimate, and the CI length. \textbf{remove the term
performance measure above and add that performance measures are bias in
all estimands, and cov and CIL of regression coeff.}

ANSWER

In general there is more bias in the estimate in conditions with a
higher proportion of missingness and or a lower number of iterations. As
required the values of or hat R lower for conditions with a higher
number off iterations. End somewhat higher 4 conditions with a higher
percentage of missing data. However autocorrelation did not show a
decrease with increase number of iterations. After replacing the ACF
function with a manual calculation for autocorrelation the
autocorrelation values were indeed decreasing with a higher number of
iterations. Evaluation with the performance measures shows that are had
an autocorrelation are conservative: signs of non-convergence in
conditions where estimands are unbiased and or confidence valid. As
expected complete convergence (as indicated by autocorrelation equals 0;
R hat equals 1) is not observed.

QUESTION: When?

Usually, convergence is diagnosed when Rhat \textless{} 1.2 or 1.1 or
1.01. And a t-test is performed for AC values == 0. Performance measures
are the same as above: unbiased, confidence valid estimates.

ANSWER

Our head is equal to 1.2 is not stringent enough because the performance
measures R not unbiased and confidence fellas fell it 4 every condition
in which our hat is smaller than one point 2. At least, not when our hat
is smaller than one point two fur first occurs. The threshold R hat is
equal to one point one is somewhat conservative with respect to the
performance measures. It seems OK. Our hat is equal to 1.01 is too
stringent compared to the performance measures (\textbf{but may be
necessary in more complex complete data models???}).

RECOMMENDATIONS

For empirical researchers: 1) Check trace plots for pathological non
convergence and adjust imputation model if necessary. 2) use our hat
wait 1.1 ash threshold and autocorrelation with \textbf{?} as threshold
. Keep iterating until these thresholds are reached. 3) Do not use the R
function ACF. Instead compute autocorrelations manually , see for
example GitHub link. 4) Track your own scalar summary of interest. This
is advanced but explained in Van Buren 2018. Compute R hat and
autocorrelation values for this scalar summary.

LIMITATIONS/FUTURE RESEARCH

Pmm, M(N)AR, Empirical data, Vignette, ShinyMICE, AC across imputations,
not iterations (unique for imp data??)

If I don't include it in my study, also: Significance of AC values;
Estimate (PCA or reg. coeff.) versus Rhat and AC on the estimate

\hypertarget{start-methods-section}{%
\subsection{Start methods section}\label{start-methods-section}}

\textbf{Add to this section: 1) explicit mention of simulation
conditions; 2) emphasize that the plots are averages across repetitions,
not within MICE; 3) sample effects due to single complete dataset}
Convergence of the MICE algorithm is investigated through model-based
simulation in \texttt{R} \citep[version 3.6.3;][]{R}. The simulation
set-up is summarized in the pseudo-code below. The complete \texttt{R}
script of the simulation study is available from
\href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.

@@

@@
